Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-06-21 paper/sec/intro.tex
add comment in alg 1

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} unexpectedly published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
The algorithm is displayed in its entirety in~\cref{alg:cvm}.
% Here, $\cdot \getsr \cdot$ denotes random sampling, and $\Ber(p)$ denotes a Bernoulli distribution with parameter $p$.

\begin{algorithm}[h!]
	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$ \Comment fail if $\chi$ remains full
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}

>>>>

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} unexpectedly published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
The algorithm is displayed in its entirety in~\cref{alg:cvm}; h.
% Here, $\cdot \getsr \cdot$ denotes random sampling, and $\Ber(p)$ denotes a Bernoulli distribution with parameter $p$.

\begin{algorithm}[h!]
	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment randomly sample a bit $b$ from the Bernoulli distribution
    \If{$b$} \Comment insert $a_i$ ($b$ is true with prob. $p$)
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else \Comment remove $a_i$ otherwise
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$ \Comment fail if $\chi$ remains full
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}
